---
title: "Coding Temple R-Programming: Guided Demo"
author: "Ken Wood"
date: "2025-02-16"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
#### 1. Setting up the Environment

Import necessary libraries and load the dataset:

```{r message=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(dplyr)
life_exp <- read.csv("data/Life Expectancy Data.csv")
```
#### 2. Initial Exploration of the Dataset

Explore the structure and summary of the dataset:

```{r summary}
str(life_exp)  # View the structure of the dataset
head(life_exp)  # Preview the first few rows
summary(life_exp)  # Summary statistics of the dataset
skim(life_exp)  # Detailed skim of the dataset
```
Check for missing data:

```{r missing_data}
missing_data <- colSums(is.na(life_exp))
missing_percent <- (missing_data/nrow(life_exp))*100
missing_df <- data.frame(
    variable = names(missing_data),
    missing_percent = missing_percent
)
```

Visualize missing data:

```{r}
ggplot(missing_df, aes(x = reorder(variable, missing_percent),
                    y = missing_percent)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(title = "Percentage of Missing Values by Variable",
        x = "Variables", y = "Missing Percentage")

```

#### 3. Data Cleaning and Transformation

Convert Year column to Date format, create new variables:

```{r}
life_clean <- life_exp %>%
    mutate(
        Year = as.Date(paste0(Year, "-01-01")),  # Convert Year to Date format
        # Create a Development Status variable
        Development_Status = ifelse(Status == "Developing", "Developing", "Developed"),  
        GDP_per_capita = GDP / Population  # Create new meaningful variable for GDP per capita
    ) %>%
    filter(!is.na(Life.expectancy))  # Remove rows with missing life expectancy values
```

Group by Country and Development Status to calculate summary statistics:

```{r}
country_stats <- life_clean %>%
    group_by(Country,Development_Status) %>%
    summarise(
        Avg_Life_Exp = mean(Life.expectancy),
        Avg_GDP_per_capita = mean(GDP_per_capita),
        Avg_Schooling = mean(Schooling, na.rm = TRUE),
        Avg_BMI = mean(BMI, na.rm = TRUE))  %>%
    arrange(desc(Avg_Life_Exp))
```

#### 4. Data Visualization

Time Trend Plot (Life Expectancy by Development Status)

```{r}
yearly_trends <- life_clean %>%
    group_by(Year, Development_Status) %>%
    summarise(
        Avg_Life_Exp = mean(Life.expectancy),
        Avg_GDP_per_capita = mean(GDP_per_capita),
        Avg_Alcohol = mean(Alcohol, na.rm = TRUE),
        Avg_BMI = mean(BMI, na.rm = TRUE)
    ) %>%
    ungroup()

time_plot <- ggplot(yearly_trends,
                    aes(x = Year, y = Avg_Life_Exp, color = Development_Status)) +
    geom_line(linewidth = 1) +
    geom_point() +
    theme_minimal() +
    labs(title = "Life Expectancy Trends Over Time",
        subtitle = "Comparing Developed vs Developing Countries",
        x = "Year", y = "Average Life Expectancy", color = "Development Status") +
    theme(legend.position = "bottom")
print(time_plot)
```

Scatter Plot (Life Expectancy vs GDP per Capita):

```{r}
scatter_plot <- ggplot(life_clean,
                    aes(x = GDP_per_capita, y = Life.expectancy, color = Development_Status)) +
    geom_point(alpha = 0.6) +
    scale_x_log10() +
    geom_smooth(method = "lm", se = FALSE) +
    theme_minimal() +
    labs(title = "Life Expectancy vs GDP per Capita",
        x = "GDP per Capita (log scale)", y = "Life Expectancy", color = "Development Status")
print(scatter_plot)
```

Box Plot (Life Expectancy by Development Status):

```{r}
box_plot <- ggplot(life_clean,
                aes(x = Development_Status, y = Life.expectancy, fill = Development_Status)) +
    geom_boxplot() +
    geom_jitter(alpha = 0.1) +
    theme_minimal() +
    labs(title = "Life Expectancy Distribution by Development Status",
        x = "Development Status", y = "Life Expectancy")
print(box_plot)
```

#### 5. Statistical Analysis: t-Test

- **Hypothesis:** Do developed countries have significantly higher life expectancy than developing countries?

Perform a t-test:

```{r}
t_test_result <- t.test(Life.expectancy ~ Development_Status, data = life_clean)
print(t_test_result)
```

#### 6. Correlation and Regression Analysis

Correlation Matrix:

```{r}
correlation_matrix <- life_clean %>%
    select(Life.expectancy, GDP_per_capita, Schooling, BMI) %>%
    cor(use = "complete.obs")
print(correlation_matrix)
```

Linear Regression Model: Predict life expectancy using GDP per capita, schooling, and BMI:
```{r}
model <- lm(Life.expectancy ~ GDP_per_capita + Schooling + BMI, data = life_clean)
summary(model)
```
